Univariate Poisson Model
========================================================

Let's explore the Poisson model a bit and simulate some data using ```rpois()```:
```{r}
rpois(n=100, lambda=1)
```

We can plot the "empirical frequency histogram" for simulated data with 3 different values of parameter $\lambda$:
```{r, tidy=FALSE}
  plot(c(0,35), c(0,1), type="n",
       xlab="x", ylab="Frequency")

  x <- rpois(n=100, lambda=1)
       hist(x, add=T, freq=FALSE, col="blue")
  x <- rpois(n=100, lambda=5)
       hist(x, add=T, freq=FALSE, col="red")
  x <- rpois(n=100, lambda=20)
       hist(x, add=T, freq=FALSE, col="green")
```

Now we can plot the respective **probability mass functions**:
```{r, tidy=FALSE}
  plot(0:35, dpois(0:35, lambda=1), 
       col="blue", pch=19, 
       ylab="Probability mass",
       xlab="x")
  points(0:35, dpois(0:35, lambda=5), col="red", pch=19)
  points(0:35, dpois(0:35, lambda=20), col="green", pch=19)
```
***

Poisson likelihood
------------------

I will generate the artificial data again:
```{r}
x <- rpois(n=100, lambda=8)
x
```

Poisson distribution is didactically convenient as it only has the one parameter $\lambda$. We can write the *deviance* (negative log-likelihood) function for a given dataset $x$:
```{r, tidy=FALSE}
  deviance.function <- function(x, lambda)
  {
    LL <- dpois(x, lambda, log=TRUE) # the log likelihood
    deviance <- -sum(LL)
    return(deviance)
  }
```

We can try different values of $\lambda$:
```{r}
  deviance.function(x, lambda=10)
```

And we can evaluate the ```deviance.function``` for values of $\lambda$ between 0 and 35:
```{r, tidy=FALSE}
  lambdas <- 0:35
  deviances <- numeric(36) # empty numeric vector
  
  for(i in 1:36)
  {
    deviances[i] <- deviance.function(x, lambdas[i])  
  }
```

Our maximum likelihood exstimate (MLE) of $\lambda$ is:
```{r}
MLE <- lambdas[which.min(deviances)] 
MLE
```

And this is the plot of the deviance function, together with the 
MLE (vertical line):
```{r}
plot(lambdas, deviances, ylab="Deviance (Negative Log Likelihood)")
abline(v=MLE)
```






